Extracting The Power of Dependent Types

### 2021-07-14 paper.tex
Full pass over paper, mostly cosmetic changes
Address some of my pet peeves

One powerful method for enforcing almost any functional property of a
program statically is by using \emph{dependent types}, as used for
example in Coq~\cite{Coq-8-13-2} or Agda~\cite{agda-2-6-2}.  In a
dependently typed language, properties can be encoded as types, and if
the program typechecks, the property is guaranteed to hold.  If we can
extend our existing language with dependent types (for verification
part) and then eliminate them (for using the original compiler), then
our goal is achieved.

@@

To keep presentation as simple as possible, we have
chosen a very simple target language to embed --- we use a subset
of PostScript~\citep{postscript1999language}, a stack-based
language used in the corresponding document format.
%
\figref{sqsum} shows an example of a program
written in our embedding and the PostScript code produced by
our extractor.
%
While the embedding may look simplistic, this simplicity
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We ensure that many verification patterns found in bigger
languages are covered, showing that the proposed approach scales.

This entire paper is a literate Agda script: all Agda code is typechecked
and all the examples are run while generating the paper.

@@

\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper.  This is
reassuring, as it means our proposed approach is immediately portable
into many other contexts.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and very much
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we can't do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.

@@

We apply this idea in the context of dependently-typed
language Agda.  We demonstrate the use of dependent types
in order to enforce static properties of the embedded programs.
Concretely, we have demonstrated the approach by implementing
an extractor for a fragment of the PostScript language.

>>>>

One powerful method for enforcing almost any functional property of a
program statically is by using \emph{dependent types}, as used for
example in Coq~\cite{Coq-8-13-2} or Agda~\cite{agda-2-6-2}.  In a
dependently typed language, properties can be encoded as types, and if
the program typechecks, the property is guaranteed to hold.  If we
extend our existing language with dependent types (for the verification
part) and then eliminate them (for using the original compiler), then
our goal is achieved.

@@

To keep presentation as simple as possible, we have
chosen a simple target language to embed --- we use a subset
of PostScript~\citep{postscript1999language}, a stack-based
language used in the corresponding document format.
%
\figref{sqsum} shows an example of a program
written in our embedding and the PostScript code produced by
our extractor.
%
While the embedding may look simplistic, this simplicity
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We ensure that many verification patterns found in bigger
languages are covered, showing that the proposed approach scales.

This entire paper is a literate Agda script: all code is typechecked
while generating the paper.

@@

\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper.  This is
reassuring, as it means our proposed approach is immediately portable
into many other contexts.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we cannot do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.

@@

We apply this idea in the context of dependently-typed
language Agda.  We demonstrate the use of dependent types
to enforce static properties of the embedded programs.
Concretely, we have demonstrated the approach by implementing
an extractor for a fragment of the PostScript language.
